---
title: "LECTURE 03: Basic concepts in statistics"
subtitle: "ENVS475: Exp. Design and Analysis"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, echo = FALSE, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', warning=FALSE, message=FALSE, fig.retina = 2)
library(tidyverse)
library(kableExtra)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(text_bold_color = "#000000",
  base_color = "#F03297",
  base_font_size = "28px",
  header_font_google = google_font("Josefin Sans"),
  text_font_google = google_font("Montserrat"),
  #code_font_google = google_font("Inconsolata", "500"),
  link_color = "#0A0AD3",
  inverse_background_color = "#4C4F57",
  title_slide_background_color = "#4C4F57",
  table_row_even_background_color = lighten_color("#4C4F57", 0.9)
  )
```

```{r plot theme, include=FALSE}
  options(digits = 4)
  library(ggplot2)
  theme_spacelab <- function(base_size = 12, base_family = "") {
    half_line <- base_size / 2
    theme(
      # Elements in this first block aren't used directly, but are inherited
      # by others
      line = element_line(
        size = 0.5, linetype = 1, colour = "black",
        lineend = "butt", color ="#f89406"
      ),
      rect = element_rect(fill = "white", colour = "black", size = 0.5, linetype = 1),
      text = element_text(
        family = base_family, face = "plain", colour = "black",
        size = base_size,
        hjust = 0.5, vjust = 0.5, angle = 0, lineheight = 0.9, margin = margin(),
        debug = FALSE
      ),
      axis.text = element_text(size = rel(0.8), color = "#999999"),
      axis.title = element_text(size = rel(0.8), color = "grey40", vjust = 0.35),
      strip.text = element_text(size = rel(0.8)),

      axis.line = element_line(size = 0.8, color = "#999999"),
      axis.line.x = NULL,
      axis.line.y = NULL,
      axis.text.x = element_text(
        size = base_size * 1.2, lineheight = 0.9,
        margin = margin(t = 0.8 * half_line / 2), vjust = 1
      ),
      axis.text.x.top = element_text(margin = margin(b = 0.8 * half_line / 2), vjust = 0),
      axis.text.y = element_text(
        size = base_size * 1.2, lineheight = 0.9,
        margin = margin(r = 0.8 * half_line / 2), vjust = 0.5
      ),
      axis.text.y.right = element_text(margin = margin(l = 0.8 * half_line / 2), hjust = 0),
      axis.ticks = element_line(size = 0.2, color = "#999999"),
      axis.title.x = element_text(
        size = base_size * 1.8, vjust = 0.3,
        margin = margin(t = 10, b = 0.8 * half_line / 2)
      ),
      axis.title.x.top = element_text(margin = margin(b = half_line), vjust = 0),
      axis.title.y = element_text(
        size = base_size * 1.8, angle = 90, vjust = 1,
        margin = margin(r = 10, l = 0.8 * half_line / 2)
      ),
      axis.title.y.right = element_text(angle = -90, margin = margin(l = half_line), vjust = 0),
      axis.ticks.length = grid::unit(0.3, "lines"),

      legend.background = element_rect(colour = NA),
      legend.spacing = unit(0.4, "cm"),
      legend.spacing.x = NULL,
      legend.spacing.y = NULL,
      legend.margin = margin(
        half_line, half_line, half_line,
        half_line
      ),
      legend.box.margin = margin(0, 0, 0, 0, "cm"),
      legend.box.background = element_blank(),
      legend.box.spacing = unit(0.4, "cm"),
      legend.key = element_rect(colour = "#999999"),
      legend.key.size = grid::unit(1.2, "lines"),
      legend.key.height = NULL,
      legend.key.width = NULL,
      legend.text = element_text(size = base_size, color = "#999999"),
      legend.text.align = NULL,
      legend.title = element_blank(),
      legend.title.align = NULL,
      legend.position = "right",
      legend.direction = NULL,
      legend.justification = "center",
      legend.box = NULL,

      panel.background = element_rect(colour = NA),
      panel.border = element_rect(fill = NA, color = NA, size = .5),
      panel.grid = element_blank(),
      panel.grid.minor = element_blank(),
      panel.spacing = grid::unit(half_line, "pt"),
      panel.ontop = FALSE,
      panel.spacing.x = NULL,
      panel.spacing.y = NULL,


      strip.background = element_rect(colour = NA),
      strip.text.x = element_text(size = base_size * 1.6, margin = margin(t = half_line, b = half_line), face = "bold", color = "#999999"),
      strip.text.y = element_text(size = base_size * 1.6, angle = -90, margin = margin(l = half_line, r = half_line), face = "bold", color = "#999999"),
      strip.switch.pad.grid = unit(0.1, "cm"),
      strip.switch.pad.wrap = unit(0.1, "cm"),
      strip.placement = "inside",

      plot.tag = element_text(
        size = rel(1.2), color = "#999999",
        hjust = 0.5, vjust = 0.5
      ),
      plot.tag.position = 'topleft',

      plot.background = element_rect(colour = NA),
      plot.title = element_text(color = "grey30",
        size = base_size * 2, hjust = 0, #face = "bold",
        vjust = 2, margin = margin(b = half_line * 1.2)
      ),
      plot.subtitle = element_text(color = "#999999",
        size = base_size * 1.2, hjust = 0,
        vjust = 2, margin = margin(b = half_line * 0.9)
      ),
      plot.caption = element_text(color = "#999999",
        size = rel(0.9), hjust = 1,
        vjust = 2, margin = margin(t = half_line * 0.9)
      ),
      plot.margin = margin(
        half_line, half_line, half_line,
        half_line
      ), complete = TRUE
    )
  }


theme_set(theme_spacelab())
  
update_geom_defaults("point", list(size = 3))
update_geom_defaults("line", list(size = 0.8))

```

class: inverse

# Outline<sup>1</sup>

.footnote[[1] Lecture based on [FANR6750](https://rushinglab.github.io/FANR6750/index.html)]

#### 1) What is statistics, and why do we need it?

--

#### 2) Populations and samples

--

#### 3) Summary statistics

--

#### 4) Inferential statistics

--
#### 5) Hypothesis testing

---

# What is statistics?

<br/>
<br/>
> The study of the collection, analysis, interpretation, presentation, and organization of data (Dodge 2006)  

--
<br/>
> The science of learning from data (various) 


--
$Statistics = Information + Uncertainty$

--
* Or:

--

  * Detecting the signal from the noise.  

---
# Why do we need statistics?  
- Summarize and describe data  

--
  - Summarize a Vector of data into a location and measure of dispersion  

--
- Test hypotheses  

--
  - i.e., are the means different between groups?  

--
- Causal inference  

--
  - Does $x$ influence $y$?  

--
- Estimate parameters  

--
- Make predictions that account for uncertainty  


---
# Populations Vs. Samples  
#### Population  
- A collection of subjects of interest  
- Often, a biologically meaningful unit  
- Sometimes a process of interest  

--

#### Sample  
- A finite subset of the population  
  + i.e. the data we collect  
- Samples allow us to draw inferences about the population  
- Good samples are:  
  + Random, Representative, Sufficiently large  

---
# Population: Human Height  

<u>P</u>opulation <u>P</u>arameters
- The "true" mean and standard deviation  

```{r pop, fig.width=9, fig.height=4}
# Assumed average American Male height = 69 \pm 2.5 inches
# Assumed average American Female height = 63 \pm 2.5 inches
x <- seq(from = 55, to = 77, length.out = 100)

pop_height <- data.frame(x = x, 
                       height = c(dnorm(x, 69, 2.5),
                                  dnorm(x, 63, 2.5)),
                       sex = rep(c("Male", "Female"), each = 100))

ggplot(pop_height) +
  geom_path(aes(x = x, y = height, color = sex), size = 1.75) +
  scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Distribution of Height in the US") +
 # theme_bw() +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))
```


---
# Sample: Human Height  
```{r samp1, fig.width=12, fig.height=6}
# Assumed average American Male height = 69 \pm 2.5 inches
# Assumed average American Female height = 63 \pm 2.5 inches
x <- seq(from = 55, to = 77, length.out = 100)

set.seed(333)
n_samp = 20
samp_height <- data.frame( 
                       height = c(rnorm(n_samp, 69, 2.5),
                                  rnorm(n_samp, 63, 2.5)),
                       sex = rep(c("Male", "Female"), each = n_samp))

mean_df <- samp_height %>%
  group_by(sex) |>
  summarise(mean_height = mean(height))


ggplot() +
  geom_histogram(data = samp_height,
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.4,
                 position = "dodge") +
  geom_path(data = pop_height, 
            aes(x = x, y = height, color = sex),
            size = 1.75) +
  geom_rug(data = samp_height,
           aes(x = height, color = sex),
           size = 1) +
  geom_vline(data = mean_df,
             aes(xintercept = mean_height, color = sex),
             size = 2) +
  
  scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Sample of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))+
  scale_fill_manual(values = c("darkorchid3", "darkorange2"))
```




---
# Sample: Human Height  
```{r samp1 annotated, fig.width=12, fig.height=6}
ggplot() +
  geom_histogram(data = samp_height,
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.4,
                 position = "dodge") +
  geom_path(data = pop_height, 
            aes(x = x, y = height, color = sex),
            size = 1.75) +
  geom_rug(data = samp_height,
           aes(x = height, color = sex),
           size = 1) +
    scale_x_continuous("Height") +
  scale_y_continuous("") +
  geom_vline(data = mean_df,
               aes(xintercept = mean_height, color = sex),
             size = 2) +
  labs(title = "Sample of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))+
  scale_fill_manual(values = c("darkorchid3", "darkorange2")) +
  annotate("text", x = 75, y = 0.2, label = "Populations",
           size = 10) +
  geom_segment(aes(x = 75, y = 0.18, xend = 64.5, yend = 0.14),
               arrow = arrow(length = unit(0.25, "cm")),
               linewidth = 0.75) +
  geom_segment(aes(x = 75, y = 0.18, xend = 70.5, yend = 0.14),
               arrow = arrow(length = unit(0.25, "cm")),
               linewidth = 0.75) +
  annotate("text", x = 72.5, y = 0.27, label = "Sample means",
           size = 10) +
  geom_segment(aes(x = 72.5, y = 0.25, xend = 62.9, yend = 0.2),
               arrow = arrow(length = unit(0.25, "cm")),
               linewidth = 0.75) +
  geom_segment(aes(x = 72.5, y = 0.25, xend = 68.7, yend = 0.2),
               arrow = arrow(length = unit(0.25, "cm")),
              linewidth = 0.75) +
  annotate("text", x = 57, y = 0.18, label = "Sample \nfrequencies",
           size = 10) +
  geom_segment(aes(x = 57, y = 0.15, xend = 62, yend = 0.037),
               arrow = arrow(length = unit(0.25, "cm")),
              linewidth = 0.75) +
  geom_segment(aes(x = 57, y = 0.15, xend = 67.9, yend = 0.05),
               arrow = arrow(length = unit(0.25, "cm")),
              linewidth = 0.75)
```




---
# Summary Statistics  
### Describing the sample  
.pull-left[
Vector of data is hard to interpret  
```{r}
set.seed(123)
tibble(Heights = round(rnorm(5, 69, 3.5), 1)) %>%
  kable("html", align = 'c') %>%
  kable_styling(bootstrap_options = c("condensed"), 
                full_width = FALSE, font_size = 18)
```
]

--

.pull-right[
Describe any vector with two summary stats  
* Average (mean)  
* Standard Deviation  
]

---
# Summary Statistics  
.pull-left[
* **Mean** is a measure of location or central tendency  
  * Works well with normally distributed data  
  * "Peak" of the bell curve  

* **Standard deviation (SD)** measure of dispersion  
  * Average distance of data point to mean  
  * "Width" of the bell curve  
]

.pull-right[
```{r, fig.width=8, fig.height=6}
ggplot(pop_height) +
  geom_path(aes(x = x, y = height, color = sex), size = 1.75) +
  scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Distribution of Height in the US") +
 # theme_bw() +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))
```
]
---
# Sample Variation 
.pull-left[
```{r samp var, fig.width=9, fig.height=3}
set.seed(333)
n_samp = 5
samp1 <- data.frame( 
                       height = c(rnorm(n_samp, 69, 2.5),
                                  rnorm(n_samp, 63, 2.5)),
                       sex = rep(c("Male", "Female"), each = n_samp))

samp2 <- data.frame( 
                       height = c(rnorm(n_samp, 69, 2.5),
                                  rnorm(n_samp, 63, 2.5)),
                       sex = rep(c("Male", "Female"), each = n_samp))

ggplot() +
  geom_histogram(data = samp1,
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.6,
                 position = "dodge") +
geom_path(data = pop_height, 
            aes(x = x, y = height, color = sex),
            size = 1.75) +
  geom_rug(data = samp1,
           aes(x = height, color = sex),
           size = 1) +
    scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Sample 1 of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))+
  scale_fill_manual(values = c("darkorchid3", "darkorange2"))
```

```{r samp var2, fig.width=9, fig.height=3}
ggplot() +
  geom_histogram(data = samp2,
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.6,
                 position = "dodge") +
geom_path(data = pop_height, 
            aes(x = x, y = height, color = sex),
            size = 1.75) +
  geom_rug(data = samp2,
           aes(x = height, color = sex),
           size = 1) +
    scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Sample 2 of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_color_manual(values = c("darkorchid3", "darkorange2"))+
  scale_fill_manual(values = c("darkorchid3", "darkorange2"))
```
]
--
.pull-right[
<br/> 

* Each sample produces its own mean and SD 

* Inferential statistics allows us to deal with this  
]

---
# Sampling = Uncertainty

Populations (usually) cannot be measured  
* sampling is essential  

--

But sampling is inherently *stochastic* 

- sampling produces uncertainty

- unavoidable (but that's ok!) 

---
# Inferential Statistics  
Standard Error (SE) of the mean  
* On average, how far is sample mean, $\bar y$, from the population mean, $\mu$?  

Statistics is what allows us to learn about the **population** using **samples** in the face of **uncertainty** 

Use samples to make *inferences* about the populations   
---
# Hypothesis Testing  

Statistics is often concerned with testing hypotheses  
* Is the sample mean different or the same as some value?  
* Are the sample means of two groups the same or different?  

We will go through the details in future classes  

The following is a brief review of statistical terms and principles  

---
# Statistics review  

What is a Null and Alternative Hypothesis?  
Can you write out one set of two-tailed hypotheses?  
Do you remember what a one-tailed hypothesis is?  

What does $\alpha$ mean?  
What value does $\alpha$ usually have?  

What is a $\text{p-values}$? 
What do we compare $\text{p-values}$ to?  

How do we decide if we reject or fail to Reject a null hypothesis?

---
# Statistics review  

Null Hypothesis = base-level we're testing against.  
  * Usually has the $=$ sign in it.  
  * i.e., Not different  

--

Alternative Hypothesis   
  * Does not have an $=$ sign in it  
  * In this class, it will have the $\ne$ sign in it.  

--

Two-tailed hypotheses
Null: $\large H_0 : \mu_1 = \mu_2$  
Alternative: $\large H_0 : \mu_1 \ne \mu_2$  
* In this class, we will always be working with 2-tailed tests  

---
# Statistics review  
$\alpha$ = level of signifigance  
  * In this class, $\alpha = 0.05$.  
  
--

$\text{p-values}$ = One of the outputs of most statistical tests  
* We compare the $\text{p-values}$ against $\alpha = 0.05$  
* If $\text{p-value} < \alpha$, we **Reject the null hypothesis **  
  + i.e., This usually means there *is a difference*  
  
-- 

* If $\text{p-value} > \alpha$, we **FAIL to reject the null hypothesis **  

  + i.e., This usually means there *is NOT a difference*  
  
---
# Example  

Is the average height of males and females in the US different?

```{r, fig.width=12, fig.height=3}
ggplot(data = samp_height) +
  geom_histogram(
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.9,
                 position = "dodge") +
    scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Sample of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_fill_manual(values = c("darkorchid3", "darkorange2"))
```

--

* Estimated difference in heights = 5.9 inches  

--

* $\text{p-value} = 0.001$  
--

What do we conclude?  

---
# Example 2  

Is the average height of males and females in the US different?

```{r, fig.width=12, fig.height=3}
n_samp = 20
samp_height2 <- data.frame( 
                       height = c(rnorm(n_samp, 67, 5),
                                  rnorm(n_samp, 65, 5)),
                       sex = rep(c("Male", "Female"), each = n_samp))
ggplot(data = samp_height2) +
  geom_histogram(
                 aes(x = height,
                     y = ..density..,
                     fill = sex), 
                 color = "white",
                 alpha = 0.9,
                 position = "dodge") +
    scale_x_continuous("Height") +
  scale_y_continuous("") +
  labs(title = "Sample of Height in the US") +
  #theme_bw(base_size = 12) +
  scale_fill_manual(values = c("darkorchid3", "darkorange2"))
```
--

* Estimated difference in heights = 1.5 inches  

--

* $\text{p-value} = 0.273$  

--

What do we conclude?  

---
# Looking ahead

#### **Wednesday**: Normal Distributions  
* Look over Normal Distribution Lecture notes before next class!  

#### **Friday**: HW02  

<br/>

#### **Reading**: Hector chp. 5

#### **Video Resources**: on D2L
